Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-12-01 leadership_across_cultures_contexts.Rmd
Working on MS, all sections.
Updated variables definition tables.

Across human societies, the influence of leaders stands to substantially shape the nature of group structure and group outcomes. Simultaneously, social norms, pressures from competing leaders or rival or cooperative groups, and followers preferences can greatly constrain a leader's agenda. The interactions between the strategies of individual leaders, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural evolutionary processes [@henrich_big_2015].

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and at all levels of social organization. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have also been relatively under-investigated. Current outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is shaped by population-level variation and the costs and benefits leaders and followers incur across distinct leadership systems.

@@

We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study, before presenting results and interpretations. Through novel empirical and exploratory analyses encompassing the full range of cultural diversity and social contexts, we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

@@

Leaders emerge in all human societies. That is, within every population there are particular contexts in which one individual maintains a disproportionate level of influence in the group and functions to facilitate group goals and provides services to the group. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @lewis_leaders_1974; @von_rueden_leadership_2015].

Despite the universality of the phenomenon of leadership, the nature of leadership across populations and across social contexts is highly diverse and the emergence of leaders and the functions they serve are closely tied to group structure. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders tends to emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, specialized skills or knowledge, and social support to achieve or maintain community influence [@kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of collective actions conflict resolution [@glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies, at the community level, representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of leaders. Within communities, however, leaders may emerge whenever individuals for groups. The role of leaders in ritual contexts, such as in initiations, and in providing social services, including facilitating marriages and organizing social activities, are also expected to be common functions leaders serve.

Increased sedentarization and defensibility of resources is often associated with both social stratification and institutionalized managerial roles including stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. In the wake of economic intensification, the functions of leadership becomes increasingly associated with the managing economic systems and distribution of resources. Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on dispersed, less defensible resources [@lopez_evolutionary_2017].

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service and nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions associated with their role. The structure of such organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

Scholars in managerial studies have long discussed the behaviors of leaders and managers have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [@bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leadership functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, creating structures for accomplishing group goals, defending institutional integrity, and maintain internal consistency.

@wofford_factor_1970 provides an early empirical view of managerial functions. A factor analysis of managerial behaviors from over 130 US employees' responses to 183 questionnaire items identified five interpretable, orthogonal factors: (1) an *order and group achievement* factor, identifying participative, democratic, and organizational planning function; (2) a *personal enhancement oriented* factor, associated with authority and control, (3) a *personal interaction* factor, associated with interpersonal relationships and employee support, (4) a *security and maintenance* factor, associated with self-preservation, and (5) a *dynamic and achievement oriented* factor, associated with aggressive, self-assured actions. Wofford's analysis provided a foundation for future empirical studies of leadership and management functions. He contrasted factor one type management, described as the professional administrator who effectively collaborates and integrates team members, from factors two and five, described as autocratic leaders strongly motivated by personal power and willing to rely on aggressive tactics [@wofford_factor_1970].

More recently, many managerial scholars have adopted a comparative approach to better understand cultural diversity in leadership systems, drawing on Culturally Endorsed Implicit Leadership Theory (CLT), which suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles and behaviors [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leadership effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied of industries across 24 countries. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, leaders moderate their behavior through culturally learned values and leadership-specific values.

@@

Across culturally and economically diverse populations there are some broad commonalities in the qualities of community leaders. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have differential influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have increased influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

Across disciplines many theoretical models and empirical findings have implicated high intelligence as an important quality positively associated with effective leadership [@johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009]. Recent cross-cultural evidence and from a small-scale society implicates strong relationships between perceptions of intelligence and leadership among males [@garfield_evolutionary_2019; @garfield_investigating_2019], when considered in the context of broad interdisciplinary results suggests high intelligence and specialized knowledge are universal traits of political and community leaders. Recent evidence from within managerial studies, however, suggests a curvilinear effect in that exceptionally intelligence individuals at high end of the distribution may not be the most effective leaders in contemporary business settings [@antonakis_can_2017].

Within managerial studies scholars often suggest the cognitive demands of leadership escalate with increasing organizational complexity [@zaccaro_organizational_2002]. More complex organizations require leaders equipped to process complex information, understand distinct departmentalized bodies, and resolve conflicting points of view, as well as be equipped to socially manage diverse individuals [@bass_handbook_1990; @schneider_organizations_2006; @antonakis_does_2009].

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners [@garfield_evolutionary_2019; @pietraszewski_evolution_2019; @hagen_leadership_2019].

<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formalized institutionalized roles and informal ephemeral roles, are also often if not universally associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional, but leadership roles through their association with increased social status are widely associated with numerous social, material, and reproductive benefits [@kantner_identifying_2010].

<!-- BENEFITS: SOCIAL BENEFITS, SOCIAL NETWORK -->
Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many industries and organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].

<!-- NEED BETTER CITATIONS ABOVE -->

<!-- BENFITS: MATING -->
Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel metaanalysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIST -->
@von_rueden_leadership_2015 review a large body of theory and ethnography and outline four possible mechanisms by which leaders in small-scale societies might offset the potential costs associated with various leadership roles; leaders may, (1) facilitate collective actions that yield net benefits to themselves or their close kin; (2) claim a fee for their services, such a tax or greater share of returns; (3) receive reciprocal exchange in other currencies from followers who have benefited from their services; or (4) may gain other social or reproductive benefits by signaling their high quality as a mate and social partner.

<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models of leadership generally assume the benefits of leadership outweigh the costs, for both leaders and followers. However, the unique costs associated with various leadership positions have received less attention, relative to the benefits of leadership, across disciplines. Group responsibilities necessarily impose opportunity costs on individual priorities and agendas, given leadership responsibilities require investments of time, energy, and resources that could be invested elsewhere.

<!-- COSTS: INCREASED SOCIAL OR PHYSICAL RISK FROM MEDIATION - SSS -->
Mediating conflicts between individuals can be physically and socially costly [@wiessner_power_2010]. Devising and proposing compromises and optimal outcomes can be cognitively challenging and the solutions leaders propose are unlikely to be equally satisfactory for everyone involved. Therefore, leaders stand to become the targets of animosity and social ridicule. In the absence of state-based systems and institutionalized inequalities, all group members are on relative equal footing and equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including overt physical aggression [@boehm_purposive_2008]. Physically formidable individuals with strong social networks are likely to be able o bear these costs more-so than their counterparts and unsurprisingly larger body size and high social capital are qualities often associated with leaders in small-scale societies [@glowacki_leadership_2015]. With greater population density and wealth inequality, these risks are expected to concomitantly increase in magnitude. In managing collective action and enacting group-level policies, some behaviors of leaders will upset some followers and leader-follower interactions can be hostile.

<!-- COSTS: INCREASED SOCIAL OR PHYSICAL RISK FROM MEDIATION - LSS -->
Conceptualizing leadership as a possible solution to problems associated with collective goods [@olson_logic_1965], where group members pool resources and share costs of public goods to receive a net benefit from a cooperative endeavor, leader incur costs associated with managing contributions, coordinating behaviors, and broadcasting information to maintain coordination and cooperation. In this framework, leaders and followers voluntary participate in a cooperative enterprise with the expectation that benefits will outweigh the imposed costs. @rich_dynamics_1980 adopts this framework of collective costs and benefits to understand leadership and group dynamics in urban neighborhoods and neighborhood organizations, which can be viewed as self-interested groups which face group-wide problems and necessarily share public benefits. The costs associated with leadership in neighborhood organizations follow those associated with other collective enterprises (e.g., investments of time and energy in management and maintenance of group coordination). @rich_dynamics_1980 conceptualizes four categories of the benefits associated with neighborhood organization. Leaders of neighborhood organizations (1) may receive benefits of the public goods produces from collective actions (as other non-leaders would as well); (2), may receive selective benefits from surpluses produced through collective actions; (3) may receive incidental benefits from their leadership role; and (4) leaders of neighborhood organizations may enjoy their service in the position. @rich_dynamics_1980 interviewed 54 officers from 11 metropolitan neighborhood organizations in the Midwest US. Results from extensive interviews suggest personal fulfillment and enjoyment are the main motivational drivers of participation in leadership roles of neighborhood organizations, suggesting the process of leadership for individuals is distinct from theoretical models rooted in economics. Therefore, a primary benefit of leadership, in some contexts, is likely to be intrinsic personal fulfillment and the opportunity to execute personal values of community service.

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations. Here, we provide the first broad and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample. We also provide the first comparative empirical view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how the qualities and functions of leaders vary across social contexts and levels of cultural complexity.

>>>>

Across human societies, the influence of leaders stands to substantially shape the nature of group structure and group outcomes. Simultaneously, social norms, competing leaders, pressures from rival or cooperative groups, and followers preferences can greatly constrain a leader's agenda. The interactions between the strategies of individual leaders, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural evolutionary processes [@henrich_big_2015].

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and at all levels of social organization. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have been relatively under-investigated. Current outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is shaped by population-level variation at different levels of social organizations and the costs and benefits leaders and followers incur across distinct leadership systems.

@@

We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study, before presenting results and interpretations. Through novel exploratory analyses encompassing the full range of cultural diversity and social contexts, we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

@@

Leaders emerge in all human societies. That is, within every population there are particular contexts in which one individual maintains a disproportionate level of influence in the group and functions to facilitate group goals and provides services to the group. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

Despite the universality of the phenomenon of leadership, the nature of leadership across populations and across social contexts is highly diverse and the emergence of leaders and the functions they serve are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, specialized skills or knowledge, and social support to achieve or maintain community influence [@kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies, at the community level, representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of leaders. Within communities, however, leaders may emerge whenever individuals form groups. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services, including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions leaders serve.

Increased sedentarization and defensibility of resources is often associated with both social stratification and institutionalized managerial roles including stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources. Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on dispersed, less defensible resources [@lopez_evolutionary_2017].

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service, nonprofit leaders, and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions associated with their roles. The structure of such organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

Scholars in managerial studies have long discussed the behaviors of leaders and managers have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [see @bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and often overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leadership functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, create structures for accomplishing group goals, defend institutional integrity, and maintain internal consistency among group members.

@wofford_factor_1970 provides an early empirical view of managerial functions. A factor analysis of managerial behaviors from over 130 US employees' survey responses identified five interpretable, orthogonal factors: (1) an *order and group achievement* factor, identifying participative, democratic, and organizational planning function; (2) a *personal enhancement oriented* factor, associated with authority and control, (3) a *personal interaction* factor, associated with interpersonal relationships and employee support, (4) a *security and maintenance* factor, associated with self-preservation, and (5) a *dynamic and achievement oriented* factor, associated with aggressive, self-assured actions. Wofford's analysis provided a foundation for future empirical studies of leadership and management functions. He contrasted factor one type management, described as the "professional administrator" who effectively collaborates and integrates team members, from factors two and five, described as "autocratic leaders" strongly motivated by personal power and a willingness to rely on aggressive tactics [@wofford_factor_1970].

More recently, many managerial scholars have adopted a comparative approach to better understand cultural diversity in leadership systems, drawing on Culturally Endorsed Implicit Leadership Theory (CLT), which suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied of industries across 24 countries. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, leaders moderate their behavior through culturally learned values and leadership-specific values.

@@

Across culturally and economically diverse populations there are some broad commonalities in the qualities of community leaders. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

Across disciplines many theoretical models and empirical findings have implicated high intelligence as an important quality positively associated with effective leadership [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009]. Recent cross-cultural evidence and from a small-scale society implicates strong relationships between perceptions of intelligence and leadership among males [@garfield_evolutionary_2019; @garfield_investigating_2019]. These results suggests high intelligence and specialized knowledge are universal traits of political and community leaders.

Within managerial studies scholars often suggest the cognitive demands of leadership escalate with increasing organizational complexity [@zaccaro_organizational_2002]. More complex organizations require leaders equipped to process complex information, understand distinct departmentalized bodies, and resolve conflicting points of view, as well as be equipped to socially manage diverse individuals [@bass_handbook_1990; @schneider_organizations_2006; @antonakis_does_2009]. However, recent evidence suggests a curvilinear effect in that exceptionally intelligence individuals at the high end of the distribution may not be the most effective leaders in contemporary business settings [@antonakis_can_2017].

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019]. 
<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formalized institutionalized roles and informal ephemeral roles, are also often if not universally associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional, but leadership roles through their association with increased social status are widely associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. 
<!-- BENEFITS: SOCIAL BENEFITS, SOCIAL NETWORK -->
Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many industries and organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].
<!-- NEED BETTER CITATIONS ABOVE -->
<!-- BENFITS: MATING -->
Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel metaanalysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIST -->
@von_rueden_leadership_2015 review a large body of theory and ethnography and outline four possible mechanisms by which leaders in small-scale societies might offset the potential costs associated with various leadership roles; leaders may, (1) facilitate collective actions that yield net benefits to themselves or their close kin; (2) claim a fee for their services, such a tax or greater share of returns; (3) receive reciprocal exchange in other currencies from followers who have benefited from their services; or (4) gain other social or reproductive benefits by signaling their high quality as a mate and social partner.

<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models of leadership generally assume the benefits of leadership outweigh the costs, for both leaders and followers. However, the unique costs associated with various leadership positions have received less attention, relative to the benefits of leadership, across disciplines. Group responsibilities necessarily impose opportunity costs on individual priorities, given leadership responsibilities require investments of time, energy, and resources that could be invested elsewhere. 
<!-- COSTS: INCREASED SOCIAL OR PHYSICAL RISK FROM MEDIATION - SSS -->
Mediating conflicts between individuals can be physically and socially costly [@wiessner_power_2010]. Devising and proposing compromises and optimal outcomes can be cognitively challenging and the solutions leaders propose are unlikely to be equally satisfactory for everyone involved. Therefore, leaders stand to become the targets of animosity. In the absence of state-based systems and institutionalized inequalities, all group members are on relative equal footing and equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including overt physical aggression or indirect aggression, such as gossip and social ridicule [@boehm_purposive_2008; @hess_gossip_2017]. Physically formidable individuals with strong social networks are likely to be able to bear these costs more so than their counterparts. Larger body size and high social capital are qualities often associated with leaders in small-scale societies [@glowacki_leadership_2015]. With greater population density and wealth inequality, these risks are expected to concomitantly increase in magnitude. In managing collective action and enacting group-level policies, some behaviors of leaders will upset some followers and leader-follower interactions can be hostile.

<!-- COSTS: INCREASED SOCIAL OR PHYSICAL RISK FROM MEDIATION - LSS -->
Conceptualizing leadership as a possible solution to problems associated with collective goods [e.g., @olson_logic_1965; @glowacki_leadership_2015; @hooper_theory_2010; @pietraszewski_evolution_2019], where group members pool resources and share costs of public goods to receive a net benefit from a cooperative endeavor, leaders incur costs associated with managing contributions, coordinating behaviors, and broadcasting information to maintain coordination and cooperation. In this framework, leaders and followers voluntary participate in a cooperative enterprise with the expectation that benefits will outweigh the associated costs. @rich_dynamics_1980 adopts this framework of collective costs and benefits to understand leadership and group dynamics in urban neighborhoods and neighborhood organizations in WEIRD settings, which can be viewed as self-interested groups that face group-wide problems and necessarily share public benefits. The costs associated with leadership in neighborhood organizations follow those associated with other collective enterprises (e.g., investments of time and energy in management and maintenance of group coordination). @rich_dynamics_1980 conceptualizes four categories of the benefits associated with neighborhood organization. Leaders of neighborhood organizations (1) may receive benefits of the public goods produces from collective actions (as other non-leaders would as well); (2), may receive selective benefits from surpluses produced through collective actions; (3) may receive incidental benefits from their leadership role; and (4) leaders of neighborhood organizations may enjoy their service in the position. @rich_dynamics_1980 interviewed 54 officers from 11 metropolitan neighborhood organizations in the Midwest US. Results from extensive interviews suggest personal fulfillment and enjoyment are the main motivational drivers of participation in leadership roles of neighborhood organizations, suggesting the process of leadership for individuals is distinct from theoretical models rooted in economics. Therefore, a primary benefit of leadership, in some contexts, is likely to be intrinsic personal fulfillment and the opportunity to execute personal values of community service.

NEED CONCLUSTION STATEMENT/PARAGRAPH.

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. Here, we seek to provide the first empirical and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how the qualities and functions of leaders vary across social contexts and levels of cultural complexity.
